### 2025-04-07 code/analysis-code/statistical-analysis.qmd
additions to statistical analysis

---
title: 
author: 
date: 
output: html_document
---
# loading in packages
```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
```
#Load in the data
```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

@@

>>>>

---
title: 
author: 
date: 
output: html_document
---
# loading in packages
```{r}
library(tidyverse)
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(ranger)

```
#Load in the data
```{r}
#Path to data. 
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

#More advanced temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
#For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.

```{r}
mydata <- mydata %>%
  drop_na()%>%
  mutate(
    `Create date` = as.Date(`Create date`),
    year = year(`Create date`),
    day_of_year= yday(`Create date`),
    week = week(`Create date`),
    month = month(`Create date`),
    season = case_when(
      month %in% c(12, 1, 2)  ~ "Winter",
      month %in% c(3, 4, 5)   ~ "Spring",
      month %in% c(6, 7, 8)   ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    `Isolation source` = factor(`Isolation source`)
  )
```

#Feature selection
```{r}
model1_data <- mydata %>%
  select(`Isolation type`,year,month,day_of_year,week,season)
```

#Split the data
```{r}
set.seed(1234)
data_split <- initial_split(model1_data, prop = 0.8, strata = `Isolation type`)
train_data <-training(data_split)
test_data <-testing(data_split)
```
#Create recipe
```{r}
rf_recipe <- recipe(`Isolation type` ~ month + year + day_of_year + week + season, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

#Define the model
```{r}
rf_mod <-rand_forest(trees = 150, min_n =5) %>%
  set_mode("classification") %>%
  set_engine("ranger")
```
Create workflow
```{r}
workflow1 <- workflow() %>%
  add_model(rf_mod)%>%
  add_recipe(rf_recipe)
```
#fit the random forest model
```{r}
fit1 <- workflow1 %>%
  fit(data = train_data)
```
#Make predictions for evaluation
```{r}
predictions_rf <-fit1 %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data)

```
```{r}
#set isolation source as a factor
predictions_rf <- predictions_rf %>%
  mutate(`Isolation type` = as.factor(`Isolation type`))

@@

#Run metrics to evaluate the model
```{r}

# Create a metric set for classification
classification_metrics <- metric_set(accuracy, precision, recall, f_meas)

# Now use it on your predictions
metrics_rf <- predictions_rf %>%
  classification_metrics(truth = `Isolation type`, estimate = .pred_class)

```{r}
metrics_rf <- predictions_rf %>%
  metrics(truth = `Isolation type`, estimate = .pred_class)
print(metrics_rf)
```
#Plot the results of this model
#Clean up some of the changes made to the data and add to processing file
